---
title: "Bioinformatics Lab 5"
author: |
  | Biswas Kumar - Bisku859
  | Karthikeyan Devarajan  - karde799
date: "11/25/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GEOquery)
library(affy)
library(RColorBrewer)
library(affyPLM)
library(limma)
library(annotate)
library(ggplot2)
library(seriation)
library(plotly)
```
  
# Question 1  
*Go to the webpage http://snap.stanford.edu/biodata/ and choose one of the provided datasets. Download it and reproduce the statistics concerning the graph. If you obtain different values, then discuss this in your report. Visualize the graph. The next step is to try to identify some clusters (communities in the graph). You can follow the tutorial at https://psych-networks.com/r-tutorial-identify-communities-items-networks/ to achieve this. Once you have found some clusters, identify the elements in it and try to find information on this cluster. Is it related to some known biological phenomena? If you do not find anything, then document your search attempts. If it will not be possible to do this question on the whole downloaded graph, then you may take some sub-graph of it.*  

```{r}
test <- read.table(file=file.choose())
```
  

# Question 2  
*Recreate one of the three analyses that can be found on https://strimmerlab.github.io/ software/genenet/index.html. Document and discuss all your steps. In the analyses there is the step where you select the edges to keep. There a particular criterion is chosen for edge inclusion. Vary this criterion and explore how the resulting clusters will differ with the changes. Take one found cluster, identify the elements in it and try to find information on this cluster. Is it related to some known biological phenomena? If you do not find anything, then document your search attempts.*    

```{r}
p <- plot_ly(x=colnames(scaleData), y=rownames(scaleData), z = scaleData, type = "heatmap")
p
```


